---
title: "Introduction to theft"
author: "Trent Henderson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Introduction to theft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 7,
  fig.width = 7,
  warning = FALSE,
  fig.align = "center"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(theft)
```

## Purpose

`theft` enables the standardised calculation of time-series features from multiple existing feature sets, and any user-supplied features.

## Core functionality

To explore package functionality, we are going to use a dataset that comes standard with `theft` called `simData`. This dataset contains a collection of randomly generated time series for six different types of processes. The dataset can be accessed via:
  
```{r, message = FALSE, warning = FALSE, eval = FALSE}
theft::simData
```

The data follows the following structure:
  
```{r, message = FALSE, warning = FALSE}
head(simData)
```

### Calculating time-series features

The core function in `theft` is `calculate_features`. You can choose which subset of features to calculate with the `feature_set` argument. The choices are currently `"catch22"`, `"feasts"`, `"tsfeatures"`, `"tsfresh"`, `"tsfel"`, and `"kats"`.

Note that `kats`, `tsfresh` and `tsfel` are Python packages. The R package `reticulate` is used to call Python code that uses these packages and applies it within the broader *tidy* data philosophy embodied by `theft`. `theft` currently provides access to $>1100$ features from these six sets alone. However, as discussed in the functionality demonstrations below, you can also supply your own list of features too!

#### Installing Python feature sets

Prior to using `theft` (only if you want to use the `Kats`, `tsfresh` or `TSFEL` feature sets; the R-based sets will run fine) you should have a working Python 3.9 installation and run the function `install_python_pkgs(venv, python)` after first installing `theft`, where the `venv` argument is the name of the virtual environment you want to create and `python` is the path to the Python interpreter you want to use.

For example, if you wanted to install the Python libraries to the default virtual environment folder used by `reticulate`, you would run the following after first having installed `theft` (here I am just creating a new virtual environment called `"theft-package"`---you can call it whatever you like!):

```{r, eval = FALSE}
install_python_pkgs(venv = "theft-package", python = "/usr/local/bin/python3.10")
```

You can then run the following to activate the virtual environment:

```{r, eval = FALSE}
init_theft("theft-package")
```

You are now ready to commit theft!

**NOTE 1: You only need to call ** `init_theft` **once per session.**

**NOTE 2: There are also separate installation functions for each Python feature set, such as** `install_tsfresh` **if you only need one of the libraries and want to keep your dependencies light.**

#### Calculating features

You are then ready to use the rest of the package's functionality, beginning with the extraction of time-series features. Here is an example with the `catch22` set on a dataset of simulated time series that ships with `theft` consisting of data from a Gaussian Noise process and an AR(1) process:
  
```{r, message = FALSE, warning = FALSE}
feature_matrix <- calculate_features(data = simData, feature_set = "catch22")
head(feature_matrix)
```

Note that `data` must be a `tsibble::tbl_ts` object, which has specified `key` (i.e., identifying) and `index` (i.e., time) variables. `theft` treats the first variable in the `key` as the ID variable and the second as the grouping variable (if there is one). Any other key variables will be ignored by `theftdlc` functions which operate only on this information.

NOTE: If using the `tsfresh` feature set, you might want to consider the `tsfresh_cleanup` argument to `calculate_features`. This argument defaults to `FALSE` and specifies whether to use the in-built `tsfresh` relevant feature filter or not.

You can also supply your own named list of functions to compute as time-series features. Below is an example with mean and standard deviation. Note that the list *must* be named as `theft` uses the list element names to label the time-series features internally. Note that if you don't want to use any of the existing feature sets in `theft` and only calculate the features you supply to `features`, just set `feature_set = NULL`.

```{r, message = FALSE, warning = FALSE}
feature_matrix2 <- calculate_features(data = simData, 
                                      feature_set = NULL,
                                      features = list("mean" = mean, "sd" = sd))

head(feature_matrix2)
```

### Comparison of feature sets

For a detailed comparison of the six feature sets, see [this paper](https://ieeexplore.ieee.org/document/9679937) for a detailed review^[T. Henderson and B. D. Fulcher, "An Empirical Evaluation of Time-Series Feature Sets," 2021 International Conference on Data Mining Workshops (ICDMW), 2021, pp. 1032-1038, doi: 10.1109/ICDMW53433.2021.00134.].

## Reading and processing hctsa-formatted files

As `theft` is based on the foundations laid by [`hctsa`](https://github.com/benfulcher/hctsa), there is also functionality for reading in `hctsa`-formatted Matlab files and automatically processing them into tidy dataframes ready for feature extraction in `theft`. The `process_hctsa_file` function takes a string filepath to the Matlab file and does all the work for you, returning a dataframe with naming conventions consistent with other `theft` functionality. As per `hctsa` specifications for [Input File Format 1](https://time-series-features.gitbook.io/hctsa-manual/installing-and-using-hctsa/calculating/input_files#input-file-format-1-.mat-file), this file should have 3 variables with the following exact names: `timeSeriesData`, `labels`, and `keywords`. The filepath can be a local drive path or a URL.

## Analysing, interpreting, and visualising time-series features

Please see the companion package [`theftdlc`](https://github.com/hendersontrent/theftdlc) ('`theft` downloadable content') for a large suite of functions.
